---
title: "2022-09-05_TidyTuesday"
author: "Douglas Lawton"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

## TidyTuesday: September 06, 2022
### LEGO sets

#### Load in libraries

```{r}
library(tidyverse)
library(mgcv)
library(DHARMa)
library(gratia)
library(patchwork)
library(knitr)
library(kableExtra)
```

#### Read in data

```{r}
inventories <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/inventories.csv.gz')
inventory_sets <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/inventory_sets.csv.gz')
sets <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/sets.csv.gz')
```

#### Cleaning the data

```{r}

all_df <- left_join(inventories, inventory_sets, by = "set_num") |>
  left_join(sets, by = "set_num") 

ex_plot <- all_df |> 
  ggplot(aes(x = num_parts)) +
  geom_density() +
  scale_x_log10()

ex_plot
```


#### Lets explore how the LEGO sets have changed through the yeats.

Perhaps set complexity increased with years?

```{r}
all_df %>%
  group_by(year) %>%
  summarize(mean_parts = mean(num_parts)) %>%
  mutate(mean_parts=round(mean_parts,2)) %>%
  arrange(-mean_parts) 
  
```

Looks like there might be something to my idea? Lets graph it by boxplots to see

```{r}

plot <- all_df %>%
  ggplot(aes(x=year,y=num_parts)) +
    geom_point(size=0.25,alpha=0.3) +
    geom_smooth(method="gam",se=F) +
    geom_smooth(method="lm",se=F,color="red") +
    scale_y_continuous(trans="log10")

plot

```

What are these 1 piece sets?

```{r}
all_df %>% filter(num_parts == 1)
```

It appears to be special situations like the whole bioncle line up -- lets remove them to get a more telling story here. Im just going to give a coarse cut off -- there needs to be at least 20 pieces in the set.

```{r}
plot2 <- all_df %>%
  filter(num_parts >= 100) %>%
  ggplot(aes(x=year,y=num_parts)) +
    geom_point(size=0.25,alpha=0.3) +
    geom_smooth(method="gam",se=F) +
    geom_smooth(method="lm",se=F,color="red") +
    scale_y_continuous(trans="log10")

plot2
```

seems to be a lot of variation still but theres some positice correlation. Maybe a nonlinear one! Lets go ahead and test for that.

#### Building the GAM

```{r}
mod_data <- all_df %>%
  filter(num_parts >= 100) %>%
  mutate(version = factor(version),
         num_parts_scaled = scale(num_parts)[,1])


mod <- gam(num_parts_scaled~s(year,k=50,bs="tp") + s(version,bs="re"),select=TRUE,family = tw(link="log"),data=mod_data)
```


```{r}
summary(mod)
draw(mod)
```

```{r}

all_df %>%
  filter(num_parts >= 50) %>%
  group_by(name) %>%
  tally() %>%
  arrange(-n)

```

