---
title: "2022-08-31_TidyTuesday"
author: "Douglas Lawton"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

## TidyTuesday: August 31, 2022
### Pell grant data

#### Load in libraries

```{r}
library(tidyverse)
library(tidymodels)
library(patchwork)
library(knitr)
library(kableExtra)
```


Lets download in the data and make sure it all looks correct

```{r}
pell <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-08-30/pell.csv')

```

```{r}
pell %>% head()
```

I've copied and pasted the column names here for ease of use:

| variable  | class   | description                              |
|-----------|---------|------------------------------------------|
| STATE     | integer | State shotcode                           |
| AWARD     | double  | Award amount in USD                      |
| RECIPIENT | double  | Total number of recipients by year, name |
| NAME      | integer | Name of college/university               |
| SESSION   | integer | Session group                            |
| YEAR      | integer | Year                                     |

How many years of data do we have for each state? Looks like it starts in 1999

```{r}

pell %>% 
  select(!c(NAME,AWARD,RECIPIENT)) %>%
  distinct() %>%
  group_by(STATE) %>%
  tally() %>% filter(n<19)

```

So it looks like we have complete data for all states! 19 years of data spanning 1999 to 2018.

So this is still a lot of data, I think im interested in comparing the many public universities of North Carolina.There are 16 of them. Sadly, the data is a bit of a mess with many naming errors. Here is some quite ugly code that amends everything.


```{r}
nc_pub_universities <- c("Appalachian State University","North Carolina Agricultural and Technical State University","North Carolina School of the Arts",
                         "North Carolina Central University","North Carolina State University","Winston-Salem State University",
                         "North Carolina Agricultural and   Technical State Unive","North Carolina State Univ-Raleigh",
                         "North Carolina A&t State Univ", "East Carolina University","Elizabeth City State University","Fayetteville State University","Western Carolina University")


pell_filtered <- pell %>% 
  filter(STATE == "NC" &
         NAME != "Queens University of Charlotte" &
         NAME != "University of Mount Olive" &
         NAME != "Sherrill's University of Barber & Cosmet" &
         NAME != "Miami International University of Art &" &
         str_detect(NAME,"University of") |
         NAME %in% nc_pub_universities)

pell_filtered$NAME[pell_filtered$NAME=="University of n c - Greensboro"] <- "University of North Carolina - Greensboro"
pell_filtered$NAME[pell_filtered$NAME=="University of n c-Chapel Hill"] <- "University of North Carolina - Chapel Hill"
pell_filtered$NAME[pell_filtered$NAME=="University of n c-Wilmington"] <- "University of North Carolina - Wilmington"
pell_filtered$NAME[pell_filtered$NAME=="University of North Carolina at Asheville"] <- "University of North Carolina - Asheville"
pell_filtered$NAME[pell_filtered$NAME=="University of North Carolina at Pembroke"] <- "University of North Carolina - Pembroke"
pell_filtered$NAME[pell_filtered$NAME=="University of North Carolina Wilmington"] <- "University of North Carolina - Wilmington"
pell_filtered$NAME[pell_filtered$NAME=="University of North Carolina at Wilmington (the)"] <- "University of North Carolina - Wilmington"
pell_filtered$NAME[pell_filtered$NAME=="University of North Carolina - Chapel Hill-"] <- "University of North Carolina - Chapel Hill"
pell_filtered$NAME[pell_filtered$NAME=="North Carolina A&t State Univ"] <- "North Carolina Agricultural and Technical State University"
pell_filtered$NAME[pell_filtered$NAME=="North Carolina State Univ-Raleigh"] <- "North Carolina State University"
pell_filtered$NAME[pell_filtered$NAME=="University of Nc at Asheville"] <- "University of North Carolina - Asheville"
pell_filtered$NAME[pell_filtered$NAME=="University of Nc at Charlotte"] <- "University of North Carolina - Charlotte"
pell_filtered$NAME[pell_filtered$NAME=="University of North Carolina School of the Arts"] <- "North Carolina School of the Arts"
pell_filtered$NAME[pell_filtered$NAME=="Univ of North Carolina at Pembroke"] <- "University of North Carolina - Pembroke"

pell_filtered <- pell_filtered %>%
  mutate(NAME = gsub("University of North Carolina","UNC",NAME),
         NAME = gsub("North Carolina","NC",NAME)) 


# Looking at all the available data for the NC public universities

pell_filtered %>%
  group_by(NAME) %>%
  count()

```

# Data exploration

Alright now with all the wrangling done, lets vizualize how pell grant awards have changed over time. There are still a bunch of schools, so I am going to chunk it out into three groups: top, middle, and bottom pell grant earners in 2018 just to help with visualization. Of course a big issue is that some of the schools are really big (e.g. NC state with about 30k students) while others are small (UNC Asheville is about 3k students).

```{r,fig.align='center'}



quantile_2018 <- pell_filtered %>%
    filter(YEAR == 2017)  %>%
    mutate(quantile = ntile(AWARD,3)) %>%
    select(quantile,NAME) %>%
    distinct()

quantiles <-pell_filtered %>%
  left_join(quantile_2018,by="NAME") %>%
  drop_na(quantile) %>%
  mutate(avg_pell_grant = AWARD/RECIPIENT) %>%
  group_split(quantile)



p <- lapply(quantiles, function(d) ggplot(data = d,aes(x=YEAR,y=avg_pell_grant,color=NAME)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels=scales::dollar) +
  scale_color_viridis_d() +
  xlab("") + ylab("") +
  ggpubr::theme_pubr(base_size=8) +
  theme(legend.title=element_blank(),legend.position = "right"))

wrap_plots(p,ncol=1)

```
 


What about overall -- how has the UNC system faired?

```{r}

pell_filtered %>%
  mutate(avg_pell_grant = AWARD/RECIPIENT) %>%
  ggplot(aes(x=YEAR,y=avg_pell_grant)) +
  geom_line(aes(group=NAME),size=0.75,alpha=0.25) +
  geom_smooth(se=FALSE,size=2,method="gam") +
  ggpubr::theme_pubr() +
  ylab("Average pell grant size")


```

Its interesting to see that huge bump around 2008 which corresponds with the great recession. 

# Normalizing by uni size

Okay, i really want to normalize this by how many students are enrolled. I went and downloaded enrollment data from the UNC website [here](https://myinsight.northcarolina.edu/t/Public/views/db_enroll/EnrollmentbyInstitution?%3Aembed=y&%3AisGuestRedirectFromVizportal=y&iid=1). Now im going to read in a relate the new data to the pell grant data.

Im going to shorten the university names too

```{r}
enrollment <- read_csv("data/unc_enrollment.csv") %>%
  select(1,2,6) %>%
  mutate(SNAPSHOT_TERM = gsub("Fall ","",SNAPSHOT_TERM),
         SNAPSHOT_TERM = as.integer(SNAPSHOT_TERM)) %>%
  rename(year ="SNAPSHOT_TERM",
         university = "Institution")

pell_filtered2 %>% filter(str_detect(name,"Pembroke"))

pell_filtered2 <- pell_filtered %>%
  select(2:6,AWARD) %>%
  mutate(university = case_when(
    NAME == "Appalachian State University" ~ "ASU",
    NAME == "East Carolina University" ~ "ECU",
    NAME == "Elizabeth City State University" ~ "ECSU",
    NAME == "Fayetteville State University" ~ "FSU",
    NAME == "NC Agricultural and Technical State University" ~ "NCA&T",
    NAME == "NC Central University" ~ "NCCU",   
    NAME == "NC School of the Arts" ~ "UNCSA",
    NAME == "NC State University" ~ "NC State",
    NAME == "UNC - Chapel Hill" ~ "UNC-CH",
    NAME == "UNC - Charlotte" ~ "UNCC",
    NAME == "UNC - Greensboro" ~ "UNCG",
    NAME == "UNC - Asheville" ~ "UNCA",  
    NAME == "UNC - Pembroke" ~ "UNCP",
    NAME == "UNC - Wilmington" ~ "UNCW",
    NAME == "Western Carolina University" ~ "WCU",  
    NAME == "Winston-Salem State University" ~ "WSSU"
  )) %>%
  #select(!NAME) %>%
  janitor::clean_names() %>%
  left_join(enrollment,by=c("university","year"))


```

```{r}
pell_filtered2 %>%
  filter(year >= 2012) %>%
  janitor::clean_names() %>%
  mutate(award_per_student = award/enrollment_measure_value) %>%
  ggplot(aes(x=year,y=award_per_student)) +
  geom_point(aes(group=university),size=0.75,alpha=0.25) +
  geom_line(aes(group=university),size=0.75,alpha=0.25) +
  geom_smooth(se=FALSE,size=2) +
  ggpubr::theme_pubr() +
  ylab("Pell Grant")
```

Interesting -- things seem to flatten out. Lets just tabulate the data and show the change

```{r}
pell_filtered2 %>%
  filter(year >= 2012) %>%
  janitor::clean_names() %>%
  mutate(award_per_student = award/enrollment_measure_value) %>%
  select(university,year,award_per_student) %>%
  pivot_wider(names_from = year,values_from =award_per_student ) %>%
  arrange(-`2017`) %>%
  mutate(across(where(is.numeric), round, 0)) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))


```





